Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-21 paper/paper.tex
one last tweak
remove comma for dangling clause

In this work, we present the design of \theoryname (\theoryabbv), a
domain-specific language for formal verification of parsers. A key
property is that \theoryabbv is an \emph{extensible} framework for
verification of parsers in that it supports the definition of grammar
formalisms of unrestricted complexity. That is, \theoryabbv is not a
system for verifying \emph{one} type of grammar formalism, but instead
is a domain-specific language in which many grammar formalisms and
their verified parsers can be implemented. For example, \theoryabbv itself is
not a verified parser generator compiling regular expressions to
deterministic finite automata, but is instead a domain specific
language in which we can write such a verified parser generator.

@@

\paragraph{Dependent Linear Types}
Our syntax for non-commutative dependent linear type theory is based on the
dependent commutative linear type theory of
\citet{krishnaswami_integrating_2015}, itself an extension of Benton's
linear-non-linear calculus \cite{benton1994}. A distinct feature of these
systems is that linear types and non-linear types are distinct sorts, so the
linear logic $!$ operator is not a primitive, but is instead definable, for
instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.  V\'ak\'ar
develops a dependent linear type theory where instead the non-linear types are
accessed using the $!$ modality, similar to Girard's original approach
\cite{vakar2015,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
categorical semantics for their system, whereas we have only developed a single
denotational model. It appears straightforward to adapt V\'ak\'ar's general
semantics approach to a non-commutative variant that would apply to
\theoryabbv. This may have applications in finding alternative models, or
developing logical relations proofs categorically.

>>>>

In this work, we present the design of \theoryname (\theoryabbv), a
domain-specific language for formal verification of parsers. A key
property is that \theoryabbv is an \emph{extensible} framework for
verification of parsers in that it supports the definition of grammar
formalisms of unrestricted complexity. That is, \theoryabbv is not a
system for verifying \emph{one} type of grammar formalism, but instead
is a domain-specific language in which many grammar formalisms and
their verified parsers can be implemented. For example, \theoryabbv itself is
not a verified parser generator compiling regular expressions to
deterministic finite automata but is instead a domain specific
language in which we can write such a verified parser generator.

@@

\paragraph{Dependent Linear Types}
Our syntax for non-commutative dependent linear type theory is based on the
dependent commutative linear type theory of
\citet{krishnaswami_integrating_2015}, itself an extension of Benton's
linear-non-linear calculus \cite{benton1994}. A distinct feature of these
systems is that linear types and non-linear types are distinct sorts, so the
linear logic $!$ operator is not a primitive, but is instead definable, for
instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.  V\'ak\'ar
develops a dependent linear type theory where instead the non-linear types are
accessed using the $!$ modality, similar to Girard's original approach
\cite{vakar2015,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
categorical semantics for their system, whereas we have only developed a single
denotational model. It should be straightforward to adapt V\'ak\'ar's general
semantics approach to a non-commutative variant that would apply to
\theoryabbv. This may have applications in finding alternative models, or
developing logical relations proofs categorically.
